### 2023-07-17 proofs/tex/alg-neg-subtyping-lemmas.tex
WIP: completeness of typing

\begin{lemma}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
        Then $[[ uN ]] = [[ ↑uP ]]$, since
        the substitution $[[ [uσ]uN ]]$ must preserve the 
        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
        and the difference in the induction statement for the positive and 
        the negative case here does not matter.

>>>>

\begin{lemma}[Completeness of the Negative Subtyping]
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
\end{lemma}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
    \begin{caseof}
        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
        Then $[[ uN ]] = [[ ↑uP ]]$, since
        the substitution $[[ [uσ]uN ]]$ must preserve the 
        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
        and the difference in the induction statement for the positive and 
        the negative case here does not matter.

### 2023-07-17 proofs/tex/constraint-merge-lemmas.tex
WIP: completeness of typing

\begin{lemma} [Completeness of Constraint Merge] 
    \label{lemma:merge-completeness}
    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$, 
    \begin{enumerate}
        \item $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined,
        \item $[[Θ ⊢ SC]]$, and
        \item $[[Θ ⊢ uσ : SC]]$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    By  definition, $[[SC1 & SC2]]$ is a union of
    \begin{enumerate}
        \item entries of $[[SC1]]$, which do not have matching entries in $[[SC2]]$,
        \item entries of $[[SC2]]$, which do not have matching entries in $[[SC1]]$, and 
        \item the merge of matching entries.
    \end{enumerate}

    This way, to show that $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, we need to demonstrate that each of these components is defined.
    To prove $[[Θ ⊢ uσ : SC]]$ we need to show that for every entry $[[scE]]$ of $[[SC]]$ restricting a variable $[[α̂±]]$,
    $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ holds.

    It is clear that the first two components of this union exist, and that if $[[scE]]$ is an entry 
    restricting $[[α̂±]]$ $[[dom(SC1) \ dom(SC2)]]$ then then $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ because  
    $[[Θ ⊢ uσ : SC1]]$ and $[[scE]] \in [[SC1]]$. Analogously, if $[[scE]]$ is an entry
    from the second component of the union, the property holds for it.  Let us show that the third component exists, and each of its entries satisfies the property.  Let us take two entries $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$ restricting the same variable $[[α̂±]]$.  $[[Θ ⊢ uσ : SC1]]$ means that $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ and $[[Θ ⊢ uσ : SC2]]$ means $[[Θ(α̂±) ⊢ [uσ]α̂± : scE2]]$.
    Then by \cref{lemma:entry-merge-completeness}, $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ is defined and $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$.
\end{proof}

>>>>

\begin{lemma} [Completeness of Constraint Merge] 
    \label{lemma:merge-completeness}
    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$, 
    $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
\end{lemma}
\begin{proof}
    By  definition, $[[SC1 & SC2]]$ is a union of
    \begin{enumerate}
        \item entries of $[[SC1]]$, which do not have matching entries in $[[SC2]]$,
        \item entries of $[[SC2]]$, which do not have matching entries in $[[SC1]]$, and 
        \item the merge of matching entries.
    \end{enumerate}

    This way, to show that $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, we need to demonstrate that each of these components is defined.

    It is clear that the first two components of this union exist. 
    Let us show that the third component exists.  
    Let us take two entries $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$ restricting the same variable $[[α̂±]]$.  $[[Θ ⊢ uσ : SC1]]$ means that $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ and $[[Θ ⊢ uσ : SC2]]$ means $[[Θ(α̂±) ⊢ [uσ]α̂± : scE2]]$.
    Then by \cref{lemma:entry-merge-completeness}, $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ is defined and $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$.
\end{proof}

### 2023-07-17 proofs/tex/decl-typing-lemmas.tex
WIP: completeness of typing

\begin{definition}[Size of a Judgement]
    \label{def:decl-typing-size}
    For a declarative typing judgement $J$
    let us define a metrics $\size{J}$ as a pair of numbers 
    in the following way:
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
            (\size{[[args]]}, \npq{[[iN]]})$)
    \end{itemize}
    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
    syntax tree of the term $[[v]]$ or $[[c]]$
    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
\end{definition}

@@

\begin{definition}[Metric]
    For a tree $T$ inferring
    a declarative typing judgement $J$,
    let us define a metric $\metric{T}$
    as a pair $(\size{J}, \eqNodes{T})$.
\end{definition}

\begin{lemma}[Declarative typing is preserved under context equivalence]
    Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
    \begin{itemize}
        \item [$+$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$.
        \item [$-$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$.
        \item [$\bullet$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us prove it by induction on the $\metric{T_1}$.
    Let us consider the last rule applied in $T_1$ (i.e., its root node).
    \begin{caseof}
        \item \ruleref{\ottdruleDTVarLabel}\\
            Then we are proving 
            that $[[Γ ; Φ1 ⊢ x : iP]]$ implies $[[Γ ; Φ2 ⊢ x : iP]]$.
            By inversion, $[[x : iP ∊ Φ1]]$, and 
            since $[[Γ ⊢ Φ1 ≈ Φ2]]$, $[[x : iP' ∊ Φ2]]$ for some $[[iP']]$ 
            such that $[[Γ ⊢ iP ≈ iP']]$.
            Then we infer $[[Γ ; Φ2 ⊢ x : iP']]$ by \ruleref{\ottdruleDTVarLabel},
            and next, $[[Γ ; Φ2 ⊢ x : iP]]$ by \ruleref{\ottdruleDTPEquivLabel}.

@@

            By inversion, we have $[[σ]]$ such that $[[Γ ⊢ σ : pas]]$ and
            $[[Γ ; Φ1 ⊢ [σ]iN ● args ⇒> iM]]$ is inferred.
            Let us denote the inference tree as $T_1'$.
            Notice that the induction hypothesis is applicable to $T_1'$:
            $\metric{T_1'} = ((\size{[[args]]}, 0), x)$ (for some $x$) is less than 
            $\metric{T_1} = ((\size{[[args]]}, |[[pas]]|), y)$ (for some $y$),
            since $|[[pas]]| > 0$ by inversion.

            This way, by the induction hypothesis, 
            there exists a tree $T_2'$ inferring
            $[[Γ ; Φ2 ⊢ [σ]iN ● args ⇒> iM]]$.
            Notice that the premises $[[args ≠ ·]]$, $[[Γ ⊢ σ : pas]]$,
            and $[[pas ≠ ·]]$ do not depend on the variable context,
            and hold by inversion.
            Then we infer $[[Γ ; Φ2 ⊢ ∀pas.iN ● args ⇒> iM]]$ by \ruleref{\ottdruleDTForallAppLabel}.
    \end{caseof}

\end{proof}

>>>>

\begin{definition}[Size of a Declarative Judgement]
    \label{def:decl-typing-size}
    For a declarative typing judgement $J$
    let us define a metrics $\size{J}$ as a pair of numbers 
    in the following way:
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
            (\size{[[args]]}, \npq{[[iN]]})$)
    \end{itemize}
    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
    syntax tree of the term $[[v]]$ or $[[c]]$
    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
\end{definition}

@@

\begin{definition}[Metric]
    \label{def:decl-typing-metric}
    For a tree $T$ inferring
    a declarative typing judgement $J$,
    let us define a metric $\metric{T}$
    as a pair $(\size{J}, \eqNodes{T})$.
\end{definition}

\begin{lemma}[Declarative typing is preserved under context equivalence]
    \label{lemma:decl-typing-context-equiv}
    Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
    \begin{itemize}
        \item [$+$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$.
        \item [$-$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$.
        \item [$\bullet$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us prove it by induction on the $\metric{T_1}$.
    Let us consider the last rule applied in $T_1$ (i.e., its root node).
    \begin{caseof}
        \item \ruleref{\ottdruleDTVarLabel}\\
            Then we are proving 
            that $[[Γ ; Φ1 ⊢ x : iP]]$ implies $[[Γ ; Φ2 ⊢ x : iP]]$.
            By inversion, $[[x : iP ∊ Φ1]]$, and 
            since $[[Γ ⊢ Φ1 ≈ Φ2]]$, $[[x : iP' ∊ Φ2]]$ for some $[[iP']]$ 
            such that $[[Γ ⊢ iP ≈ iP']]$.
            Then we infer $[[Γ ; Φ2 ⊢ x : iP']]$ by \ruleref{\ottdruleDTVarLabel},
            and next, $[[Γ ; Φ2 ⊢ x : iP]]$ by \ruleref{\ottdruleDTPEquivLabel}.

@@

            By inversion, we have $[[σ]]$ such that $[[Γ ⊢ σ : pas]]$ and
            $[[Γ ; Φ1 ⊢ [σ]iN ● args ⇒> iM]]$ is inferred.
            Let us denote the inference tree as $T_1'$.
            Notice that the induction hypothesis is applicable to $T_1'$:
            $\metric{T_1'} = ((\size{[[args]]}, 0), x)$ is less than 
            $\metric{T_1} = ((\size{[[args]]}, |[[pas]]|), y)$ for any $x$ and $y$,
            since $|[[pas]]| > 0$ by inversion.

            This way, by the induction hypothesis, 
            there exists a tree $T_2'$ inferring
            $[[Γ ; Φ2 ⊢ [σ]iN ● args ⇒> iM]]$.
            Notice that the premises $[[args ≠ ·]]$, $[[Γ ⊢ σ : pas]]$,
            and $[[pas ≠ ·]]$ do not depend on the variable context,
            and hold by inversion.
            Then we infer $[[Γ ; Φ2 ⊢ ∀pas.iN ● args ⇒> iM]]$ by \ruleref{\ottdruleDTForallAppLabel}.
    \end{caseof}
\end{proof}
